---
format:
  revealjs:
    width: 1280
    height: 720
    title-slide: false
    slide-number: true
    theme: default
    css: slides.css
    self-contained: true
    include-in-header:
      - nav-include.html
bibliography: ../gitbook/book.bib
execute:
  echo: false
  warning: false
  message: false
editor: source
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center",
                      warning = FALSE, message = FALSE)
library(ggplot2)
library(knitr)

theme_nh <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title    = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(colour = "grey40", size = base_size - 1),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

lmu_blue  <- "#00558C"
lmu_light <- "#6AB0DE"
grey50    <- "#888888"
green_ok  <- "#388E3C"
amber     <- "#F57F17"
```

##  {.cover-slide data-section="Cover"}

::::::::::::::: columns
:::::::::::: {.column width="45%"}
::::::::::: cover-left
::: cover-title
**Neural Hydrology**
:::

::: cover-subtitle
Deep Learning for Streamflow Prediction under Climate Change
:::

:::::::: cover-meta
<div><strong>Author:</strong> Yuxin Qiu</div>
<div><strong>Supervisor:</strong> Henri Funk</div>
<div><strong>Seminar:</strong> Climate Change Statistics</div>
<div><strong>University:</strong> LMU Munich</div>
<div><strong>Date:</strong> 2026-06-09 &ensp;·&ensp; Interim Meeting</div>
::::::::
:::::::::::
::::::::::::

:::: {.column width="55%"}
::: cover-right
:::
::::
:::::::::::::::

------------------------------------------------------------------------

## What is Neural Hydrology? {data-section="Introduction"}

**Streamflow** is the volume of water flowing through a river channel at a given point per unit time — typically measured in mm/day or m³/s. It is the integrated output of the entire catchment: precipitation, evaporation, soil infiltration, and groundwater dynamics all feed into it [@addor2017].

**Rainfall-runoff modelling** is the task of predicting daily streamflow $Q$ from meteorological inputs (precipitation, temperature, radiation) — a central problem in hydrology for over a century [@kratzert2018].

. . .

**Neural Hydrology** is the application of deep learning to this problem. Instead of hand-crafting physical equations, a neural network learns the input-to-streamflow mapping directly from data, across hundreds of catchments simultaneously.


## Milestones in Neural Hydrology {data-section="Introduction"}

- **2018** — First successful LSTM rainfall–runoff model, outperforming the calibrated SAC-SMA + Snow17 benchmark [@kratzert2018]
- **2019** — EA-LSTM: one model trained on 531 CAMELS basins simultaneously [@kratzert2019]
- **2021** — MC-LSTM: mass conservation enforced as a structural guarantee [@hoedt2021]
- **2022** — NeuralHydrology released as the first unified open-source deep-learning framework for hydrology [@kratzert2022joss]
- **2023** — Caravan: global large-sample hydrology benchmark with 6,830 catchments [@kratzert2023caravan]

------------------------------------------------------------------------

## Why Streamflow Prediction Matters {data-section="Introduction"}

::: {style="font-size: 1.05em;"}
-   **Floods** are the most frequent natural disaster — 57 million people affected in 2022 alone [@cred2022]
-   Accurate **streamflow forecasts** are critical for flood early-warning, reservoir management, and hydropower scheduling
-   Under **climate change**, precipitation extremes intensify, seasonal patterns shift, and snowmelt regimes change [@ipcc2021]
-   Hydrological models must be **robust to non-stationarity** — conditions outside the historical training distribution
:::

. . .

> *"Stationarity is dead: Whither water management?"* — Milly et al. (2008)

------------------------------------------------------------------------

## The Deep Learning Breakthrough {data-section="Background"}

```{r model-comparison-bar, fig.height=2.8, fig.width=8}
models <- data.frame(
  Model = factor(
    c("LSTM\n(single basin)", "SAC-SMA\n(calibrated)"),
    levels = c("SAC-SMA\n(calibrated)", "LSTM\n(single basin)")
  ),
  NSE   = c(0.63, 0.58),
  Group = c("Deep Learning", "Process-based")
)

ggplot(models, aes(x = Model, y = NSE, fill = Group)) +
  geom_col(width = 0.50) +
  geom_hline(yintercept = 0.6, linetype = "dashed", colour = grey50) +
  annotate("text", x = 0.6, y = 0.615, label = "Acceptable threshold",
           hjust = 0, size = 3.5, colour = grey50) +
  geom_text(aes(label = NSE), hjust = -0.25, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Deep Learning" = lmu_blue, "Process-based" = lmu_light)) +
  scale_y_continuous(limits = c(0, 0.80), expand = c(0, 0)) +
  coord_flip() +
  labs(x = NULL, y = "Mean NSE (241 CAMELS-US basins)",
       title = "LSTM vs. SAC-SMA+Snow-17",
       subtitle = "Experiment 1: one LSTM per basin, no manual calibration",
       fill = NULL) +
  theme_nh()
```

Process-based models (SAC-SMA+Snow-17, and others) require calibrating 10–40 parameters per catchment — not scalable to ungauged basins [@kratzert2018]. Individual LSTMs were trained separately for each of the **241 basins** with no manual calibration; mean NSE = **0.63** versus SAC-SMA mean NSE = **0.58** [@kratzert2018].

------------------------------------------------------------------------

## LSTM and EA-LSTM {data-section="Background"}

The LSTM [@hochreiter1997] maintains a **cell state** $\mathbf{c}_t$ as hydrological memory:

$$\mathbf{c}_t = \underbrace{\mathbf{f}_t \odot \mathbf{c}_{t-1}}_{\text{long-term memory}} + \underbrace{\mathbf{i}_t \odot \tanh(\mathbf{W}_g [\mathbf{h}_{t-1}, \mathbf{x}_t] + \mathbf{b}_g)}_{\text{new information}} \qquad \mathbf{h}_t = \mathbf{o}_t \odot \tanh(\mathbf{c}_t)$$

Cell state $\approx$ soil moisture / groundwater · Forget gate $\approx$ drainage · Input gate $\approx$ infiltration

. . .

**Entity-Aware LSTM** [@kratzert2019] — conditions the input gate on static catchment attributes $\mathbf{x}_s$:

$$\mathbf{i}_t = \sigma(\mathbf{W}_i \mathbf{x}_s + \mathbf{b}_i)$$

One model trained simultaneously on **531 CAMELS-US basins**: median NSE **0.72** — outperforming the best calibrated process model (SAC-SMA, NSE 0.58) by a wide margin.

The **Nash-Sutcliffe Efficiency** $\text{NSE} = 1 - \frac{\sum(\hat{q}_t - q_t)^2}{\sum(\bar{q} - q_t)^2}$: ranges from $-\infty$ to 1; above 0.6 is considered acceptable in hydrology.

------------------------------------------------------------------------

## NeuralHydrology: Python Library {data-section="Library"}

**NeuralHydrology** [@kratzert2022joss] — open-source PyTorch library built for reproducible deep learning in hydrology:

::: {style="font-size: 1.0em;"}
-   **Config-file driven** — a single `.yml` file specifies the full experiment; no source code changes between runs
-   **Model zoo** — 15+ architectures: CudaLSTM, EA-LSTM, MC-LSTM, MTS-LSTM, Transformer, Mamba, xLSTM, HybridModel
-   **Dataset zoo** — CAMELS-US/GB/DE/AUS/CL, Caravan (>6 000 basins globally)
-   **Probabilistic heads** — GMM, CMAL, UMAL for uncertainty quantification
-   **Evaluation** — NSE, KGE, MSE, hydrological signatures built-in
:::

``` bash
uv pip install neuralhydrology
nh-run train    --config-file config.yml
nh-run evaluate --run-dir    runs/my_run/
```

------------------------------------------------------------------------

## Physics and Transfer Learning {data-section="Library"}

**MC-LSTM** [@hoedt2021] — re-interprets cell states as water storage compartments. Mass conservation is a **structural guarantee**, not a loss penalty:

$$\Delta S_t = P_t - \hat{Q}_t - \text{ET}_t \quad \text{(enforced at every time step)}$$

This prevents physically impossible predictions when extrapolating to future climate forcing outside the training distribution.

. . .

**Transfer learning** [@kratzert2019] — NeuralHydrology has a dedicated fine-tuning workflow:

1. **Pre-train** on a large, diverse dataset (Caravan, 6 000+ basins) — builds a general hydrological encoder
2. **Fine-tune** on the target basin with limited data (head-only / full / layer-wise)

Pre-trained models reach higher NSE with limited target data than training from scratch — particularly useful for ungauged basins and data-scarce Alpine catchments.

------------------------------------------------------------------------

## Case Study: LamaH-CE — Danube River Network {data-section="Case Study"}

```{r lamah-results, fig.height=3.4, fig.width=8.5}
df_dan <- data.frame(
  Scenario = factor(
    c("Station 399\n(Danube near Bratislava)", "Median of 375 gauges\n(LamaH-CE baseline)", "Worst-case gauge\n(flood spike events)"),
    levels = c("Worst-case gauge\n(flood spike events)", "Median of 375 gauges\n(LamaH-CE baseline)", "Station 399\n(Danube near Bratislava)")
  ),
  NSE   = c(0.899, 0.856, 0.248),
  Group = c("Strong result", "Typical result", "Extreme event failure")
)
ggplot(df_dan, aes(x = Scenario, y = NSE, fill = Group)) +
  geom_col(width = 0.55) +
  geom_hline(yintercept = 0.6, linetype = "dashed", colour = grey50) +
  annotate("text", x = 0.65, y = 0.615, label = "Acceptable threshold (0.6)",
           hjust = 0, size = 3.5, colour = grey50) +
  geom_text(aes(label = NSE), hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Strong result"         = lmu_blue,
                               "Typical result"        = lmu_light,
                               "Extreme event failure" = amber)) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0)) +
  coord_flip() +
  labs(x = NULL, y = "NSE",
       title = "LamaH-CE: Danube river network (Central Europe), 375 gauges",
       fill = NULL) +
  theme_nh()
```

**Dataset**: LamaH-CE [@klingler2021]

------------------------------------------------------------------------

## Gap Analysis: What Remains Unsolved {data-section="Research Questions"}

```{r challenges-table}
df_ch <- data.frame(
  Challenge = c("Extreme events",
                "Long-term non-stationarity",
                "Regional transfer (Central Europe)"),
  Status = c("LSTM systematically underestimates flood peaks [@frame2022]",
             "Fine-tuning helps short-term; no systematic test under CMIP6 projections",
             "Benchmarks dominated by CAMELS-US; little validation in Alpine basins"),
  Direction = c("Probabilistic / asymmetric loss functions (GMM, UMAL)",
                "MC-LSTM vs. CudaLSTM under SSP climate forcing",
                "Caravan pre-training → GKD/LfU Bavaria fine-tune")
)
kable(df_ch,
      col.names = c("Challenge", "Current Status", "Promising Direction"),
      align = c("l", "l", "l"))
```

. . .

These three gaps define the research questions for this work.

------------------------------------------------------------------------

## RQ 1 — Probabilistic Losses for Flood-Peak Prediction {data-section="Research Questions"}

> *Can replacing the NSE loss with a probabilistic head (GMM / UMAL) improve LSTM performance on extreme high-flow events in snowmelt-dominated catchments?*

**Motivation**: @frame2022 showed that LSTM trained with NSE underestimates flood peaks because normal-flow days dominate the training gradient. Probabilistic heads model the full discharge distribution — rare peaks are no longer averaged away.

**Method**:

1. Train CudaLSTM with standard NSE loss (baseline)
2. Train with GMM and UMAL heads — both available in NeuralHydrology
3. Evaluate on peaks-over-threshold (POT) events using peak-weighted NSE and flood quantiles (Q90, Q95, Q99)

**Expected contribution**: Quantify whether distributional modelling closes the gap between LSTM skill on normal flows and on extreme events, without changing the architecture.

------------------------------------------------------------------------

## RQ 2 — Physics Constraints under Climate Forcing {data-section="Research Questions"}

> *Does MC-LSTM's structural water balance constraint yield more reliable streamflow projections than CudaLSTM under CMIP6 delta-change scenarios (SSP2-4.5 and SSP5-8.5)?*

**Motivation**: Under distributional shift (future precipitation intensities outside training range), unconstrained LSTMs may produce unphysical outputs. MC-LSTM's architectural guarantee prevents this — but the empirical magnitude of the benefit has not been systematically quantified.

**Method**:

1. Train MC-LSTM and CudaLSTM on CAMELS-US historical period (1980–2010)
2. Apply CMIP6 delta-change anomalies ($\Delta P$, $\Delta T$) to historical forcings
3. Compare projected streamflow: water balance closure error, distributional divergence from physically plausible range

**Expected contribution**: Empirical evidence for whether physics constraints provide a meaningful safety margin in climate extrapolation — informing model selection for impact studies.

------------------------------------------------------------------------

## RQ 3 — Transfer Learning to Alpine Basins {data-section="Research Questions"}

> *Can a NeuralHydrology model pre-trained on Caravan (6 000+ global basins) transfer effectively to ungauged Bavarian catchments, and which catchment attributes drive transferability?*

**Motivation**: Nearly all neural hydrology benchmarks use CAMELS-US. Alpine basins differ fundamentally: pronounced elevation gradients, glacier influence, distinct soil types, different seasonality. Whether large-sample pre-training generalises across this domain gap is open — with direct relevance for German flood early-warning.

**Method**:

1. Pre-train EA-LSTM on full Caravan dataset
2. Fine-tune on gauged Bavarian basins (GKD [@gkd2025]); evaluate on held-out basins
3. Compare to scratch training; SHAP analysis to identify which static attributes predict transferability

**Data**: Caravan [@kratzert2023caravan] · GKD Bayern [@gkd2025] · LfU flood event catalogue [@lfu2025]

**Expected contribution**: First systematic transfer learning benchmark for Alpine Germany.

------------------------------------------------------------------------

## Summary {data-section="Research Questions"}

::: {style="font-size: 1.05em; line-height: 1.9;"}
1. **LSTMs surpass process-based models** on CAMELS-US without basin-specific calibration [@kratzert2018]; EA-LSTM extends this to 531 basins simultaneously (NSE 0.72) [@kratzert2019]

2. **NeuralHydrology** provides the infrastructure — probabilistic heads, physics-constrained architectures, and Caravan support — for the proposed research [@kratzert2022joss]

3. **Three open gaps** motivate the research questions:
   - RQ 1: extreme event prediction via probabilistic losses
   - RQ 2: physics constraints under CMIP6 climate forcing
   - RQ 3: transfer learning to Alpine/Bavarian ungauged basins
:::

. . .

> *Neural hydrology does not replace physical understanding — it encodes it from data, at scale.*

------------------------------------------------------------------------

## References {data-section="References"}

::: {#refs style="font-size: 0.75em;"}
:::
